---
title: "Story 2"
author: "Andrew Bowen"
date: "2023-09-16"
output: pdf_document
---
# Story 2

```{r setup, include=FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(glue)
library(lubridate)
```

## Introduction
Inflation and its economic effects have been in the news recently. The Fed's goal has been to curb inflation and maintain low unemployment. [The Phillips Curve](https://socialsci.libretexts.org/Bookshelves/Economics/Economics_(Boundless)/23%3A_Inflation_and_Unemployment/23.1%3A_The_Relationship_Between_Inflation_and_Unemployment#:~:text=The%20Phillips%20curve%20shows%20the%20inverse%20trade%2Doff%20between%20rates,low%2C%20inflation%20will%20be%20high.) is an economic principle stating the general negative relationship between these two variables.

## Data Sources

- [New York Fed API](https://markets.newyorkfed.org/static/docs/markets-api.html#/)
- [BLS Data API](https://www.bls.gov/developers/api_signature_v2.htm#single) - used for unemployment data

The BLS API has an [R wrapper](https://www.bls.gov/developers/api_r.htm) that allows for easy data pulls from the API.


```{r}
url <- "https://api.bls.gov/publicAPI/v2/timeseries/popular?survey=AP"
response <- GET(url)

response_txt <- content(response, "text", encoding = "UTF-8")
json <- fromJSON(response_txt)

json
```
## Getting CPI time series data

```{r get cpi-time-series}
# Call CPI data from BLS
registration_key <- Sys.getenv("BLS_API_KEY")
series_id <- 'APU0000702111'
series_url <- glue("https://api.bls.gov/publicAPI/v2/timeseries/data/{series_id}?registrationkey={registration_key}")

response <- GET(series_url)
```

```{r message=FALSE}
# Format CPI data
response_txt <- content(response, "text", encoding = "UTF-8")
json <- fromJSON(response_txt, flatten = TRUE)

cpi <- as.data.frame(json[[4]])$series.data[[1]]
head(cpi)
```

We'll need to pull an additional 5 years of CPI data
```{r}
series_url <= glue("https://api.bls.gov/publicAPI/v2/timeseries/data/{series_id}?registrationkey={registration_key}&startyear=1998&endyear=2003")
response <- GET(series_url)
response_txt <- content(response, "text", encoding = "UTF-8")
json <- fromJSON(response_txt, flatten = TRUE)

cpi_older <- as.data.frame(json[[4]])$series.data[[1]]
```


```{r}
# Combining our CPI data
cpi_df <- rbind(cpi, cpi_older)

# Converting dates from BLS format
cpi_df$date <- str_c(cpi_df$periodName," ", cpi_df$year)
cpi_df$date <- myd(cpi_df$date, truncated=1)

# cast types
cpi_df$cpi <- as.double(cpi$value)
```


Now let's make a simple plot of our CPI data over time
```{r}
ggplot(cpi_df, aes(x=date, y=cpi)) + geom_line() + 
  labs(x="Month", y="Consumer Price Index (CPI)", title="Consumer Price Index over Time - All US", subtitle="The CPI has shown a steady increase over the past few years post-COVID")
```

## Grabbing Interest Rate Data

Grbabing interest rate data from the [St Louis Fed API (FRED)](https://fred.stlouisfed.org/docs/api/fred/). The [endpoint I'm using](https://fred.stlouisfed.org/series/FEDFUNDS) contains information about the Federal Effective Funds rate, which is the [interest rate charged to banks borrowing overnight to meet their reserve requirements](https://www.investopedia.com/terms/f/federalfundsrate.asp)

```{r}
# Use an environement variable for API token
fed_api_key <- Sys.getenv("FED_API_KEY")
url <- glue("https://api.stlouisfed.org/fred/series/observations?series_id=EFFR&api_key={fed_api_key}&file_type=json&frequency=m&observation_start=1998-08-01")

# Pull data from FRED API
fed_response <- GET(url)
```

Let's do some basic data wrangling of our FRED API response to get it into a good format for plotting
```{r}
# Write Response data to R dataframe
fed_response_txt <- content(fed_response, "text", encoding = "UTF-8")
fed_json <- json <- fromJSON(fed_response_txt, flatten = TRUE)
fed <- fed_json$observations
```


Let's clean up our federal funds rate data
```{r}
# Convert types
fed$date <- as.Date(fed$date)
fed$rate <- as.double(fed$value)
head(fed)
```




```{r}
# Plotting Federal Funds Rate over time
ggplot(fed, aes(x=date, y=rate)) + 
  geom_line() + labs(y="Federal Funds Rate", x="Month", title ="Effective Federal Funds Rate over Time")
```